---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(corrplot)
library(gridExtra)
```

### Load data


```{r load-data}
load("movies.Rdata")
```


* * *

## Part 1: Data

The following analysis is aimed at finding attributes that make a movie popular.  As a data scientist of Paramount picture, I am conducting a study that will be useful in order to make project regarding production of new movies.
The dataset contains information about movies released from 1970 to 2014. 
This will be an observational study due to the nature of the data (i.e. no random assignment). The study can be generalized to movies issued between 1970 and 2014.
Because the dataset is based only on reviews via rotten tomatoes and Internet Movie Database (IMDB),  it might be biased, because we are considering audience rating from only two sources.

* * *

## Part 2: Research question

Can any specific characteristic of a movie be used to predict its popularity, and if so, which of these characteristics are more useful for this prediction? This study would be useful for any media production when making decision on whether or not investing in funding a new movie.

* * *

## Part 3: Exploratory data analysis

In order to measure popularity of a movie (my dependent variable) I have three indicators in the dataset that I could use:  Rating on IMDB (imdb_rating), Critics score on Rotten Tomatoes (critics_score) and   Audience score on Rotten Tomatoes (audience_score).
I will check summary statistics of these variables and their distribution.



```{r}

summary(movies$imdb_rating)
summary(movies$critics_score)
summary(movies$audience_score)

h1<- qplot(movies$imdb_rating, geom="histogram",binwidth = 0.5, main = "Histogram of IMDB rating", xlab = "IMDB", fill=I("pink"))  

h2<- qplot(movies$critics_score, geom="histogram",binwidth = 0.5, main = "Histogram of Critics Score", xlab = "Critics", fill=I("pink")) 

h3<- qplot(movies$audience_score, geom="histogram",binwidth = 0.5, main = "Histogram of audience score", xlab = "Audience", fill=I("pink")) 

grid.arrange (h1,h2,h3, ncol=3, top="Distribution of possible dependent variables")


```

Of the three, the distribution of the IMDB variable looks more normally distributed and I will use this one to measure movie popularity.

As for independent variables, I will use only relevant variables from the data set and omit those which are not relevant to predict popularity, such as the title of the movie, url address of the review and dates of dvd release . I subset the variables that I will include in the analysis.

```{r}
movies_new <- movies %>% select(imdb_rating, title_type , genre, runtime, mpaa_rating, best_actor_win, best_actress_win, best_dir_win, thtr_rel_month)
```


Summary statistics of the variables selected in the new dataset for this analysis.
```{r}
summary(movies_new)
```


All the independent variables are categorical except runtime and month of release (thtr_rel_month), which I will convert to categorical. Runtime has also NAs. I omit the NAs and convert the month of release in a categorical variable.


```{r}
movies_new <- na.omit(movies_new)
movies_new$thtr_rel_month <- as.factor(movies_new$thtr_rel_month)
```


Graphic representation of independent variables I will use in the analyses.

```{r}
pl_genre <-ggplot(data=movies_new, aes(x= genre))+geom_bar(fill="pink")+ xlab("Movie Genre")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_title_type<-ggplot(data=movies_new, aes(x= title_type))+geom_bar(fill="pink")+ xlab("Movie Type")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_runtime<-ggplot(data=movies_new, aes(x= runtime))+geom_bar(fill="pink")+ xlab("Runtime")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_mpaa_rating <-ggplot(data=movies_new, aes(x= mpaa_rating))+geom_bar(fill="pink")+ xlab("MPAA Rating ")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_thtr_rel_month <-ggplot(data=movies_new, aes(x= thtr_rel_month))+geom_bar(fill="pink")+ xlab("Month of movie release")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_best_actor_win <-ggplot(data=movies_new, aes(x= best_actor_win))+geom_bar(fill="pink")+ xlab("Actor won an oscar")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_best_actress_win <-ggplot(data=movies_new, aes(x= best_actress_win))+geom_bar(fill="pink")+ xlab("Actress won an oscar")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
pl_best_dir_win<-ggplot(data=movies_new, aes(x= best_dir_win))+geom_bar(fill="pink")+ xlab("Director won an oscar")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0))
grid.arrange(pl_title_type, pl_genre,pl_runtime , pl_mpaa_rating, pl_thtr_rel_month, pl_best_actor_win, pl_best_actress_win, pl_best_dir_win, nrow=2, top= "Graph of dependent variables of movies")

```


From these representation we find that most of the actors and directors of the movies in the dataset never won an Oscar, which was expected because only few actors have win this award once a year. Most of the movies are released in December, January and June. This could have an effect on the rating because maybe there are certain times of the year when people are more incline to go to the movie theater. The majority of the movies in the dataset are categorized as drama and they are feature film, while only a few are Documentary and TV movies. These characteristics of the movies under analysis need to be kept into considerations because they can be source of bias. In fact the findings of the analysis might be biased for example toward drama movies because they are over represented in the sample. The analysis should therefore apply a model that consider this (i.e. by including the variable 'genre')


* * *

## Part 4: Modeling

The analysis in this project will be done through a linear regression with a model which include the variables I subset, IMDB rating and 8 dependent variables. From this full model I will check if I can create a parsimonious model with least variables with a backward elimination method based on p-value. I will therefore eliminate the one by one the variables that are not statistically significant at 0.05 level.

```{r}
full=lm(imdb_rating ~., data=movies_new)
summary(full)

```


This full model is significant as a whole, because the p-values is less than 0.05. The Adjusted R-squared:  0.2855 meaning that this model explains 28.55% of the variation in the independent variable. As opposed to the R-squared, the adjusted R-squared gives the percentage of variation explained by only those independent variables that actually affect the dependent variable. Adjusted R-squared only increases if the variables added to the model are useful in predicting the dependent variable.

The first variable I am eliminating is month of release of the movie, in fact ALL THE LEVELS of this variable are not statistically significant. If at least one level was statistically significant I could not eliminate the variable.

#MODEL1 without month of release

```{r}
model1=lm(imdb_rating ~ title_type + genre + runtime + mpaa_rating+ best_actor_win+ best_actress_win+ best_dir_win, data=movies_new)
summary(model1)

```


In this model adjusted R-squared has increased to 0.2933.

In the second model I eliminate the variable best_actor_win, whether an actor in the movie has ever won an Oscar.


#Model2 without best_actor_win
```{r}
model2=lm(imdb_rating ~ title_type + genre + runtime + mpaa_rating+ best_actress_win+ best_dir_win, data=movies_new)
summary(model2)

```


The adjusted R-squared has further increased to 0.2944.
I will create a final model eliminating also best_actress_win, whether an actress in the movie has ever won an Oscar.

#Final model without best_actress_win
```{r}
final=lm(imdb_rating ~ title_type + genre + runtime + mpaa_rating + best_dir_win, data=movies_new)
summary(final)

```


In this last model the adjusted R-squared has increased to 0.2949. In addition all the variables in the model are statistically significant at some level. We have managed to have a parsimonious model with only 5 independent variables and a higher adjusted R-squared, starting from an initial model with 8 independent variables.

#INTERPRETATIONS of the final model
The adjusted R-squared is 0.2949, meaning that the model explains 29.49% of the variability in the IMDB rating of a movie, this is not a high percentage, therefore this model has not very good predictive capability. There is about 70% unexplained variability by the model in the dependent variable.
The intercept of 6.048729 indicates the IMDB score when all independent variables in the model are zero or at reference level, which is when a movie is a documentary, in the genre of Action & Adventure, mpaa rating is G(general admission) the director did not win an Oscar, the runtime is 0. This value has no useful meaning, especially because no movie can have duration of 0. According to the model, if the director of the movie has ever won an Oscar in his career, the IMDB rating is expected to increase by 0.41 points. Also, for each additional minute of duration of the movie, the IMDB rating is expected to increase by 0.01.


#Diagnostics for model
I will check in the condition of multiple linear regression are met in my model.

1.	Linear relationship between the numerical independent variable (runtime) and the dependent variable.
```{r}
plot(final$residuals~movies_new$runtime, xlab= "Runtime", ylab="Residuals",main="Residuals vs. runtime")
```


The residuals of the model against the runtime variable are scattered around 0, therefore this condition is met.

2.	Nearly normal residuals with mean 0

```{r}
hist(final$residuals, main="Histogram for distribution of residuals")
qqnorm(final$residuals, main="Normal probability plot of residuals")
qqline(final$residuals)

```

The residuals of the final model are nearly normal distributed centered at 0, even if there is a small skew on the left, this condition is also met.

3.	Constant variability of residuals

```{r}
plot(final$residuals~final$fitted, main="Residuals vs. fitted")
plot(abs(final$residuals)~final$fitted, main="Absolute value of residuals vs. fitted")

```


The plot of residuals vs. predicted values shows that residuals are randomly scattered in a band with a constant width around 0. This condition is met.

4.	Independent residuals


I check if any time series structure is suspect.

```{r}
plot(final$residuals)
```


The plot of the residuals in the order of collection does not show any increase or decrease of residuals, therefore there is no suspicion of time series structure and we can assume independence of residuals, which basically means independent observations as well.



* * *

## Part 5: Prediction

I test my model using a movie released in 2016, Dangal. I am taking data for this movie from the IMDB website (http://www.imdb.com/title/tt5074352/), as I am using this IMDB rating as a dependent variable in my model.

```{r}
dangal<-data.frame(title_type= "Feature Film",genre="Drama", runtime=161, mpaa_rating="PG", best_dir_win="yes")
predict(final, dangal, interval = "predict")
```


The predicted rating is 7.72 and we are 95% confident that the real movie score will fall between 5.90 and 9.54. The prediction is good; in fact the actual rating is 8.6 from the IMDB website, and this value falls in the predicted interval.


* * *

## Part 6: Conclusion

The analyses of the movies database allowed us to answer my research question successfully; therefore we can conclude that we can predict the rating of a movie according to some of its known characteristic. My model explain 29.49% of the variation in the rating which is not very high, and for this reason the interval predicted will be wide. The dependent variable is the rating on the Internet movie database, therefore the model rely on the assumption that the movie is rated on this website.  This also means that the model is biased toward movies that are present on IMDB.
In order to avoid this bias we should collect data from other sources as well, for example through interviews or questionnaires over the telephone or submitted randomly via email. 
The sample is also biased toward movies produced in the USA as only a minority of movies issued other countries are shown in movie theaters  in USA and  present on IMDB and on Rotten Tomatoes. A further study could include international movies and/or international audience.  Moreover a different sample methodology could be used, for example by selecting an audience and asking them to watch the same movies and collecting feedback afterwards.

